---
title: "day_three"
author: "Dillon Niederhut"
date: "July 27, 2015"
output: pdf_document
---

# Introduction

okay, time for the business. 

analysis here is going to procede in two steps:

1. exploratory data analysis
2. statistical inference

our treatment of graphing owes a lot to the Grammar of Graphics, 

# Summarizing

## let's load in a dataset from the EdX servers about 

```{r}
dat <- read.csv()
str(dat)
```

## R provides two easy/simple summary functions in the base package

```{r}
table(dat)
summary(dat)
```

## plyr provides a few more ways to aggregate things

```{r}
```

# Plotting

## baseplot in R isn't fantastic, so we're going to use ggplot

```{r}
install.packages('ggplot2')
library(ggplot)
```

## qplot

## geom_hist

## geom_bar

## geom_point

## geom_scatter

## geom_boxplot

## stat_smooth

## titles and labels

## facetting

# Mean testing

## t.test

## aov

# Linear models

## cor.test (Pearson)

## lm

## lmtest

## robust

## MASS

# Nonparametric

## cor.test (Spearman)

## chisq

## Fisher's exact test

## permutation tests