---
title: 'Day Two: Data Cleaning'
author: ["Dillon Niederhut", "Shinhye Choi"]
date: "August, 19, 2015"
output: pdf_document
---

```{r, echo=FALSE}
knitr::opts_knit$set(root.dir = '../')
```

# Review

## Inspecting objects

we'll start by using some data that is already in R

```{r, eval=FALSE}
data(state)
str(state.x77)
```

## Inspecting variables

We should see 50 levels in this division variable

```{r}
state.division
length(state.division)
levels(state.division)
```

## Inspecting data frames

recall, a dataframe is a list of vectors, where each vector is one variable with all of its measurements

R expects dataframes to be rectangular

```{r}
state <- state.x77
rm(state.x77)
state <- as.data.frame(state)
head(state)
```

## Introduction

Today's class will be essentially be split into two components: CRUD operations in R and TIDY data. For more on tidiness in data, see [Hadley Wickham's paper](www.jstatsoft.org/v59/i10/paper). We will also touch on missingness - for an accessible introduction, you can read [this very old and no longer state-of-the-art paper](http://psycnet.apa.org/journals/met/7/2/147/).

yesterday we saw how to create dataframes in R

```{r}
my.data <- data.frame(n = c(1, 2, 3),
                      c=c('one', 'two', 'three'),
                      b=c(TRUE, TRUE, FALSE),
                      d=c(as.Date("2015-07-27"),
                          as.Date("2015-07-27")+7,
                          as.Date("2015-07-27")-7),
                      really.long.and.complicated.variable.name=999)
```

remember, you can learn about dataframes with

```{r}
str(my.data)
```

in practice, you will only rarely create dataframes by hand, because creating tables in a text editor is both boring and prone to error

## Readibility

we've broken up the previous command across multiple lines to make it easier to read
this is a stylistic choice, and one that should be encouraged: however, it won't be obvious to most of the students that it is necessary to either highlight the whole command and run, or hit run for every line, starting from the first one, in order

often, students will just run the second line, and be confused when nothing runs correctly in the console anymore - the way to get out of this is by hitting `ESC`

# Reading dataframes from file

## why read data from text files?

they are human-readable and highly interoperable

```{r}
read.table("data/mydata.csv", sep=',', header = TRUE)
```

> side note - anyone who is 100% new to computing will have a hard time understanding the concept of a working directory, and will try to run this code from their home directory (spoiler alert - it doesn't work)

## R has convenience wrappers for reading in tables

```{r}
read.csv("data/mydata.csv")
```

note that we are only reading the files by doing this

## R also has its own kind of data file

```{r}
load("data/mydata.Rda")
```

the `load` function does actually put the file into memory, and with the name you originally gave it when you saved it

this is typically a bad thing, and there is currently no easy workaround

## to read in tables from excel, use the `xlsx` package

if you are exporting data from excel, be sure to export datetimes as strings, as excel does not store dates internally the same way Unix does

```{r, eval=FALSE}
install.packages("xlsx")
library(xlsx)
read.xlsx("data/cpds_excel_new.xlsx")
```
But it may be better to save your .xlsx file as a csv. format in Excel first, and then read the csv file into R.

## you can also use R to read in data from proprietary software

```{r, eval=FALSE}
# examples of these?
install.packages("foreign")
library(foreign)
read.dta("data/cpds_stata.dta")
read.spss()
read.octave()
```

# Cleaning data

there are two major steps to data cleaning, which we will call 'sanitizing' and 'tidying'

in sanitizing, our goal is to take each variable and force its values to be honest representations of its levels

in tidying, we are arranging our data structurally such that each row contains exactly one observation, and each column contains exactly one kind of data about that observation (this is sometimes expressed in SQL terms as "An attribute must tell something about the key, the whole key, and nothing but the key, so help me Codd")

## exporting data from other software can do weird things to numbers and factors

```{r}
dirty <- read.csv('data/dirty.csv')
str(dirty)
```

## it's usually better to DISABLE R's intuition about data types

unless you already know the data is clean and has no non-factor strings in it (i.e. you are the one who created it)

```{r}
dirty <- read.csv('data/dirty.csv',stringsAsFactors = FALSE)
str(dirty)
```

## let's start by removing the empty rows and columns

```{r}
tail(dirty)
dirty <- dirty[1:5,-6]
dim(dirty)
```

## you can replace variable names

and you should, if they are uninformative or long

```{r}
names(dirty)
names(dirty) <- c("time", "height", "dept", "enroll", "birth.order")
```

## it's common for hand-coded data to have a signifier for subject-missingness

(to help differentiate it from your hand-coder forgetting to do something)

```{r}
dirty$enrollment
```

## you should replace all of these values in your dataframe with R's missingness signifier, `NA`

```{r}
table(dirty$enroll)
dirty$enroll[dirty$enroll=="999"] <- NA
table(dirty$enroll, useNA = "ifany")
```

> side note - read.table() has an option to specify field values as `NA` as soon as you import the data, but this is a BAAAAD idea because R automatically encodes blank fields as missing too, and thus you lose the ability to distinguish between user-missing and experimenter-missing


## that timestamp variable is not in a format R likes

base R doesn't handle time well, so we need to get rid of the time part of the timestamp

```{r}
dirty$time
dirty$time <- sub(' [0-9]+:[0-9]+:[0-9]+','',dirty$time)
dirty$time
```

## the height variable is in four different units

we can fix this with a somewhat complicated loop (since R started as a functional language, there are not easy ways to conditionally modify structures in place)

OR

we can do the same task line-by-line, since the number of observations is small

```{r}
class(dirty$height)
as.numeric(dirty$height)
```

because there are apostrophes and quotation marks, R thinks these are strings

```{r}
dirty$height[grep("’", dirty$height, perl=TRUE)] <- 5*30.48 + 9*2.54
dirty$height[2] <- 70*2.54
dirty$height[3] <- 2.1*100
```

## let's fix some of those department spellings

first, let's make this all lowercase

```{r}
dirty$dept
dirty$dept <- tolower(dirty$dept)
dirty$dept <- gsub(' ', '', dirty$dept)  # what did we just do?
dirty$dept[4] <- "geology"
dirty[dirty == "999"] <- NA
```

## then, you can coerce the data into the types they should be

```{r}
dirty$time <- as.Date(dirty$time,'%m/%d/%Y')
dirty$height <- as.numeric(dirty$height)
dirty$dept <- as.factor(dirty$dept)
dirty$enroll <- as.factor(dirty$enroll)
dirty$birth.order <- as.numeric(dirty$birth.order)
str(dirty)
```

# Missingness

there are many reasons why you might have missing data

*AS LONG AS MISSINGNESS IS NOT CAUSED BY YOUR INDEPENDENT VARIABLE* this is fine

deleting those observations is wasteful, but easy (listwise deletion)

ignoring the individual missing data points is typical (casewise deletion)

imputing mean values for missing data is possibly the worst thing you can do

imputing via MI + error is currently the best option

## listwise deletion is wasteful

```{r}
na.omit(dirty)
```

## casewise deletion is what R does internally

```{r}
nrow(dirty)
sum(is.na(dirty$height))
sum(is.na(dirty$birth.order))
length(lm(height ~ birth.order,data=dirty)$fitted.values)
```

this is usually the default strategy

## remember how we talked about the extensibility of R?

amelia is a package that makes a complicated MI approach work without you knowing anything about its implementation

```{r}
library(Amelia)
```

## let's use this large dataset as an example

```{r}
large <- read.csv('data/large.csv')
summary(large)
nrow(na.omit(large))
```

## for it to work you need low missingness and large N

```{r}
a <- amelia(large,m = 1)
print(a)
```

## amelia returns a list, where the first item is a list of your imputations

we only did one, so here it is

```{r}
large.imputed <- a[[1]][[1]]
summary(large.imputed)
```

## if you give it a tiny dataset, it will fuss at you

```{r}
a <- amelia(large[990:1000,],m = 1)
print(a)
```

# Reshaping

now that our data is clean, it's time to put it in a tidy format. this is a way of storing data that makes it easy to:

1. make graphs
2. run tests
3. summarize
4. transform into other formats

we are basically trying to organize ourselves such that:

1. any grouping is made on rows
2. any testing is done between columns

## an aside on testing

in R, you use double symbols for testing

```{r}
1 == 2
1 != 1
1 >= 1
```

(you've already seen a couple of these)

## tests return boolean vectors

```{r}
1 >= c(0,1,2)
```

## recall that boolean vectors need to be the same length or a divisor

if your vectors are not multiples of each other, R will fuss at you

```{r}
c(1,2) >= c(1,2,3)
c(1,2) >= c(1,2,3,4)     # why no warning this time? R recycles!
```

the combination of the length requirement, the lack of support in R for proper indexing, and missingness in your data will cause many headaches later on

## subsetting data frames

subsetting your data is where you will use this regularly

```{r}
my.data$numeric == 2
my.data[my.data$numeric == 2,]
```

## boolean variables can act as filters right out of the box

```{r}
my.data[my.data$b,]
```

you see the empty space after the comma? that tells R to grab all the columns

## you can also select columns

```{r}
my.data[,'d']
```

that empy space **before** the comma? that tells R to grab all the rows

## you can also match elements from a vector

```{r}
good.things <- c("three", "four", "five")
my.data[my.data$character %in% good.things, ]
```

## most subsetting operations on dataframes also return a dataframe

```{r}
str(my.data[!(my.data$character %in% good.things), ])
```

## subsets that are a single column return a vector

```{r}
str(my.data$numeric)
```

## most tidying can be done with two R packages

(plus a wrapper around the base string functions)

```{r, eval=FALSE}
install.packages('reshape2')
install.packages('stringr')
install.packages('plyr')
```

```{r}
library(reshape2)
library(stringr)
library(plyr)
```

## reshaping

our goal here is to arrange our data such that each table is about one kind of thing: whether it is everything about a measurement, everything about a person, or everything about a group of people

```{r}
abnormal <- data.frame(name = c('Alice','Bob','Eve'),
                       time1 = c(90,90,150),
                       time2 = c(100,95,100))
```

this table is not tidy - why not?

the table is about measurements, but each measurement does not have its own row, and each type of measurement value is represented by more than one column

```{r}
normal <- melt(data = abnormal, id.vars = 'name')
normal
```

we can `melt` this dataframe down into a long format, which makes each row a unique observation, and then clean up the dataframe a bit

```{r}
normal$id <- seq(1:nrow(normal))
names(normal) <- c('name','time','value','id')
normal$time <- str_replace(normal$time,'time','')

```

now that we are in a tidy format, see how easy it is to subset

```{r}
normal[normal$time == 1,]
normal[normal$name == 'Alice',]
```

and test

> side note - don't worry about how this works yet - we'll talk about it tomorrow

```{r}
t.test(value ~ time, data=normal)
```

it's easy to combine tidy tables to compare different levels of information simultaneously

# Merging data frames

## flexibly join dataframes with `merge`

imagine you have two datasets that you want to merge

```{r}
data.1 <- read.csv('data/merge_practice_1.csv')
data.2 <- read.csv('data/merge_practice_2.csv')
str(data.1)
str(data.2)
```

sometimes the same people have differet jobs in different locations

you can do an *inner* join using merge

```{r}
merge(data.1, data.2, by = 'id')
```

that's no good - we lost half of our people!

inner joins are mostly used when you **only** want records that appear in both tables

if you want the union, you can use an outer join

```{r}
merge(data.1, data.2, by = 'id', all = TRUE)
```

this works basically the same as `join` in SQL

running merges is particularly useful when:

a. your data is tidy; and,
b. you want to add information with a lookup table

in this case, you can store your lookup table as a dataframe, then merge it

```{r}
lookup <- read.csv('data/merge_practice_3.csv')
str(lookup)
```

this lookup table gives us the population for each location

we can add this to our people table with

```{r}
merge(data.1, lookup, by = "location")
```

note that Reno was in our lookup table

```{r}
lookup[lookup$location == 'Reno', ]
```

but doesn't show up when we merge - why do you think this is?

# Transforming data

## introduction

because R started out as a functional language, it can be hard to modify data, especially in place

in practice, if you want 100% control over how your frames are being modified, you'll be writing lots of `for` loops, which is messy

luckily, there is a package that handles the common tasks for you

```{r}
library(dplyr)
```

## sort data with `arranage`

base R syntax for sorting is a bit of a pain in that you have to create a sorting vector based on the values in a column, then subset the same dataframe and apply the sorting vector to the rows slice

to demonstrate this, let's start by making a toy data frame

```{r}
toy <- data.frame(
  id = c(1,1,1,2,2,2,3,3,3),
  score.1 = c(90,94,40,80,80,80,76,80,82)
)
arrange(toy, score.1)
```

## select rows by pattern with `select`

it's common for variables that measure similar things to have similar names, but selecting columns this was in base R requires running `grepl` on column names, then subsetting the dataframe and applying the logical vector to the column field

```{r}
toy$score.2 <- 100
select(toy, score.1, score.2)
select(toy, contains('score'))
```

## apply summary fucntions with `summarise`

dplyr includes most of the base R summary statistics, along with:

* `n()`
* `n_distinct()`
* `first()`
* `last()`

```{r}
summarise(toy, n(), n_distinct(score.1), last(score.1))
```

## dplyr allows you to apply functions to groups

so far, these have taken base R functions and made them faster (with C++ calls behind the scenes), easier to use, or both

dplyr's real utility is in its grouped dataframes, which apply dplyr functions groupwise

```{r}
group_by(toy, id)
summarise(group_by(toy, id), n(), n_distinct(score.1))
```

you can add as many functions as you want inbetween, but wrapping function call around function call can be hard to read (and write!)

## you can pipe functions with the `%>%` operator

this will look very familiar if you are used to working in bash

```{r}
toy %>% group_by(id) %>% summarise(n(), n_distinct(score.1))
```

# Practice

## Grab some data from Pew

and sanitize/tidy it

this will be hard

```{r}
library(foreign)
pew <- as.data.frame(read.spss("data/pew.sav"))
religion <- pew[c("q16", "reltrad", "income")]
rm(pew)
```

## we'll start by cleaning up the factor variables

```{r}
religion$reltrad <- as.character(religion$reltrad)
religion$reltrad <- str_replace(religion$reltrad, " Churches", "")
religion$reltrad <- str_replace(religion$reltrad, " Protestant", " Prot")
religion$reltrad[religion$q16 == " Atheist (do not believe in God) "] <- "Atheist"
religion$reltrad[religion$q16 == " Agnostic (not sure if there is a God) "] <- "Agnostic"
religion$reltrad <- str_trim(religion$reltrad)
religion$reltrad <- str_replace_all(religion$reltrad, " \\(.*?\\)", "")

religion$income <- c("Less than $10,000" = "<$10k",
  "10 to under $20,000" = "$10-20k",
  "20 to under $30,000" = "$20-30k",
  "30 to under $40,000" = "$30-40k",
  "40 to under $50,000" = "$40-50k",
  "50 to under $75,000" = "$50-75k",
  "75 to under $100,000" = "$75-100k",
  "100 to under $150,000" = "$100-150k",
  "$150,000 or more" = ">150k",
  "Don't know/Refused (VOL)" = "Don't know/refused")[religion$income]

religion$income <- factor(religion$income, levels = c("<$10k", "$10-20k", "$20-30k", "$30-40k", "$40-50k", "$50-75k",
  "$75-100k", "$100-150k", ">150k", "Don't know/refused"))
```

## now we can reduce this down to three columns for three variables

```{r}
religion <- count(religion, c("reltrad", "income"))
names(religion)[1] <- "religion"
```

# Acknowledgements

## Materials taken from:

[Chris Krogslund](https://github.com/ckrogs/r_useful_dlab)
[Hadley Wickham](http://www.jstatsoft.org/v59/i10/paper)
